Elephant Seal Age JAE Intraspecific Variation

### 2024-05-29 paper/paperv3.qmd
Fit seg mdl w dummy var, move to supp. Replace w poly2 mdl in main.

age_senesce <- 12

sealdat <- read_csv(here::here("data/raw/128L pull 2023_12_05.csv"), 
                    show_col_types = FALSE) %>% 
  mutate(observed = if_else(observed == "B", "Breeder", "Non-breeder"), 
         observed_int = if_else(observed == "Breeder", 1, 0), 
         pup_survived = pupseeneveragain > 0) %>% 
  filter(age > 3, 
         year > 1987) %>% 
  mutate(animalID = factor(animalID),
         year_fct = factor(year),
         age10 = age / 10) %>% 
  group_by(animalID) %>% 
  mutate(longevity = ifelse(max(year) < 2020,
                            max(age),
                            NA),
         longevity10 = longevity / 10) %>% 
  ungroup()

@@

# Fit separate GLMMs to the data subsetted before and after onset of senescence
# We a priori chose age 12 for that (age_senesce)
# Pre-senescence model
breed_age_mod1 <- glmer(
    observed_int ~ age10 + (1 | animalID) + (1 | year_fct),
    # Here's the age filter
    data = filter(sealdat, age < age_senesce),
    family = "binomial",
    control = glmerControl(optimizer="bobyqa")
)
# Post-senescence model
breed_age_mod2 <- glmer(
    observed_int ~ age10 + (1 | animalID) + (1 | year_fct),
    data = filter(sealdat, age >= age_senesce),
    family = "binomial",
    control = glmerControl(optimizer="bobyqa")
)

breed_summ <- tibble(
  model = c("Pre-senescence", "Post-senescence"),
  age_coef = c(fixef(breed_age_mod1)["age10"],
               fixef(breed_age_mod2)["age10"]),
  age_coef_se = c(summary(breed_age_mod1)$coef["age10", "Std. Error"],
                  summary(breed_age_mod2)$coef["age10", "Std. Error"]),
  `p(coef<0)` = c(pnorm(summary(breed_age_mod1)$coef["age10", "z value"]),
                  pnorm(summary(breed_age_mod2)$coef["age10", "z value"]))
)

breed_summ %>% 
  mutate(across(where(is.numeric), \(x) signif(x, 3))) %>%
  set_names(c("Age class", 
              "Age", 
              "se(Age)",
              "p(Age) < 0")) %>% 
  knitr::kable()

@@

# Create POPULATION-LEVEL (i.e. without rand effects) predictions for both
# models and row bind them together
breed_pred <- rbind(
  # Pre-senescence
  ggpredict(
    breed_age_mod1,
    terms = c("age10 [all]"),
    type = "fixed",
    interval = "confidence"
  ) %>% 
    as_tibble() %>% 
    mutate(age = x * 10,
           # Add an age_class variable to facilitate plotting
           age_class = "young"),
  # Post-senescence
  ggpredict(
    breed_age_mod2,
    terms = c("age10 [all]"),
    type = "fixed",
    interval = "confidence"
  ) %>% 
    as_tibble() %>% 
    mutate(age = x * 10,
           age_class = "old")
) %>% 
  filter(age <= 20)

# Now create rand effect predictions
# This utility function pulls rand effect *coefs* from a model
get_ranefs <- function(model, r, reduce = TRUE) {
  result <- ranef(model)[[r]] %>% 
    as_tibble(rownames = r) %>% 
    rename(ranef = 2) %>% 
    arrange(ranef)

  if (reduce)
    slice(result, round(seq(1, nrow(result), length.out = 5)))
  else 
    result
}

# Get year and individual random effeccts from pre- and post-senescence models
year_ranef1 <- get_ranefs(breed_age_mod1, "year_fct")
year_ranef2 <- get_ranefs(breed_age_mod2, "year_fct")
indiv_ranef1 <- get_ranefs(breed_age_mod1, "animalID")
indiv_ranef2 <- get_ranefs(breed_age_mod2, "animalID")

# This util fun pulls rand effect *predictions* from a model
get_ranef_pred <- function(model, r, keep) {
  ggpredict(model,
            terms = c("age10 [all]", r), 
                             type = "random",
                             interval = "confidence") %>% 
  as_tibble() %>% 
  mutate(age = x * 10,
         "{r}" := group) %>% 
  semi_join(keep, by = r) %>% 
  filter(age <= 20)
}

# Predictions based on 5 different *years* (min, q1, median, q3, max ranef
# coef)
breed_pred_year <- rbind(
  # Pre-senescence
  get_ranef_pred(breed_age_mod1, "year_fct", year_ranef1) %>% 
    mutate(age_class = "young"),
  # Post-senescence
  get_ranef_pred(breed_age_mod2, "year_fct", year_ranef2) %>% 
    mutate(age_class = "old")
) %>% 
  filter(age <= 20)

# As above, but 5 different *individuals*
breed_pred_indiv <- rbind(
  get_ranef_pred(breed_age_mod1, "animalID", indiv_ranef1) %>% 
    mutate(age_class = "young"),
  get_ranef_pred(breed_age_mod2, "animalID", indiv_ranef2) %>% 
    mutate(age_class = "old")
) %>% 
  filter(age <= 20)

# The actual raw data
breed_nonbreed <- sealdat %>% 
  group_by(age) %>% 
  summarize(n_observed = n(), 
            perc_breed = sum(observed_int) / n_observed, 
            se = (perc_breed * (1 - perc_breed) / n_observed)^0.5) %>% 
  filter(age <= 20)

# Figure with raw data and model predictions
breed_fig <- ggplot(breed_pred, aes(x = age)) +
  # Prediction across years (fixed effect only)
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, 
                  group = age_class),
              fill = "grey", alpha = 0.5) +
  geom_line(aes(y = predicted, color = age_class), 
            linewidth = 1.5) +
  # Predictions for specific years (random effect)
  geom_line(aes(y = predicted, group = interaction(year_fct, age_class)),
            breed_pred_year,
            linewidth = 0.75, alpha = 0.5) +
  # Raw data
  geom_pointrange(aes(y = perc_breed, 
                      ymin = perc_breed - se, 
                      ymax = perc_breed + se),
                  breed_nonbreed) + 
  # Year labels
  geom_text(aes(y = predicted, label = year_fct),
            filter(breed_pred_year, age == 20),
            x = 20.2,
            hjust = 0,
            size = 2.5) +
  geom_text(aes(y = predicted, label = year_fct),
            filter(breed_pred_year, age == 4),
            x = 3.8,
            hjust = 1,
            size = 2.5) +
  # Add visual indicator for senescence
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_x_continuous("Age", limits = c(3.5, 20.5)) +
  scale_y_continuous("Adult female breeding", 
                     limits = c(NA, 1), 
                     labels = scales::percent) +
  scale_color_manual(values = c("firebrick", "cornflowerblue")) +
  theme_few() +
  theme(legend.position = "none",
        axis.title = element_text(size = 13))

# Coefficient panel
breed_coef_fig <- breed_summ %>% 
  mutate(model = str_extract(model, ".*-"),
         model = fct_rev(model)) %>% 
  ggplot(aes(color = model)) + 
  geom_pointrange(aes(ymin = age_coef - age_coef_se,
                      y = age_coef,
                      ymax = age_coef + age_coef_se,
                      x = model)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("cornflowerblue", "firebrick")) +
  labs(x = "Senescence",
       y = "Age coefficient") +
  theme_few() +
  theme(legend.position = "none")

(breed_fig | breed_coef_fig) + 
  plot_layout(widths = c(3, 1)) +
  plot_annotation(tag_levels = "A")

@@

>>>>

age_senesce <- 11

sealdat <- read_csv(here::here("data/raw/128L pull 2023_12_05.csv"), 
                    show_col_types = FALSE) %>% 
  mutate(observed = if_else(observed == "B", "Breeder", "Non-breeder"), 
         observed_int = if_else(observed == "Breeder", 1, 0), 
         pup_survived = pupseeneveragain > 0) %>% 
  filter(age > 3, 
         year > 1987) %>% 
  mutate(animalID = factor(animalID),
         year_fct = factor(year),
         age10 = age / 10,
         age_cat = factor(age >= age_senesce, 
                          labels = c("Young", "Old"))) %>% 
  group_by(animalID) %>% 
  mutate(longevity = ifelse(max(year) < 2020,
                            max(age),
                            NA),
         longevity10 = longevity / 10) %>% 
  ungroup()

@@

breed_age_mod <- glmer(
  observed_int ~ poly(age10, 2, raw = TRUE) + (1 | animalID) + (1 | year_fct),
  sealdat,
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

summary(breed_age_mod)

@@

# Summarize the raw data
breed_nonbreed <- sealdat %>% 
  group_by(age) %>% 
  summarize(n_observed = n(), 
            perc_breed = sum(observed_int) / n_observed, 
            se = (perc_breed * (1 - perc_breed) / n_observed)^0.5) %>% 
  filter(age <= 20)

# Create POPULATION-LEVEL (i.e. without rand effects) predictions
breed_pred_pop <- ggpredict(
  breed_age_mod, 
  terms = "age10 [all]",
  type = "fixed",
  interval = "confidence"
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10) %>% 
  filter(age <= 20)

# Create WEIGHTED-AVERAGE predictions using random effects
seal_years <- count(sealdat, year_fct)
breed_pred_wgt <- ggpredict(
  breed_age_mod,
  terms = c("age10 [all]", "year_fct [all]"),
  type = "random",
  interval = "confidence"
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10,
         year_fct = group) %>% 
  filter(age <= 20) %>% 
  left_join(seal_years, by = "year_fct") %>% 
  group_by(age) %>% 
  summarize(predicted = weighted.mean(predicted, n))

ggplot(breed_pred_pop, aes(age, predicted)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), 
              alpha = 0.2) +
  geom_line(linewidth = 1.2, color = "grey50") +
  geom_line(data = breed_pred_wgt, 
            color = "cornflowerblue", 
            linewidth = 1.2) +
  geom_pointrange(aes(y = perc_breed, 
                      ymin = perc_breed - se, 
                      ymax = perc_breed + se),
                  breed_nonbreed) +
  scale_y_continuous("Breeding", labels = scales::percent) +
  labs(x = "Age (years)") +
  theme(legend.position = "none")

@@

```{r}
#| label: seg-breeding-mod

# Fit GLMMs to the data with binary variable indicating before and after onset # of senescence. We a priori chose age 11 for senescence (age_senesce).
breed_age_seg <- glmer(
  observed_int ~ age10 * age_cat + (1 | animalID) + (1 | year_fct),
  sealdat,
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

summary(breed_age_seg)
```

```{r}
#| label: fig-seg-breeding-mod

breed_pred_seg <- ggpredict(breed_age_seg, 
                            terms = c("age10 [all]", "age_cat")) %>% 
  as_tibble() %>% 
  mutate(age = x * 10,
         age_cat = factor(group, levels = c("Young", "Old"))) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce))

ggplot(breed_pred_seg, aes(age, predicted)) +
  geom_ribbon(aes(fill = age_cat, ymin = conf.low, ymax = conf.high), 
              alpha = 0.2) +
  geom_line(aes(color = age_cat)) +
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_color_manual(values = c("cornflowerblue", "firebrick")) +
  scale_fill_manual(values = c("cornflowerblue", "firebrick")) +
  scale_y_continuous("Breeding", labels = scales::percent) +
  labs(x = "Age (years)") +
  theme(legend.position = "none")
